---
title: "A.1 Midterm1 Partial Review"
subtitle: 
author: 
date: "Fall 2019"
output:
  html_document:
    df_print: paged
    toc: yes
  pdf_document:
    fig_caption: yes
    toc: yes
geometry: margin=1.5in
editor_options:
  chunk_output_type: console
---

```{r include=FALSE}
knitr::opts_chunk$set(fig.align="center", fig.height=5.5, fig.width=6, collapse=TRUE, comment="", prompt=F, echo = TRUE, cache=TRUE, autodep=TRUE, cache.comments=FALSE,
                      tidy=TRUE, tidy.opts=list(width.cutoff=60))
options(width=63)
```




# Introduction


Here we will review some of the topics/concepts that some of you find particularly difficult.  For a more extensive list of the topics that we have covered so far, see Midterm 1 Study Guide posted on Piazza.


# Exercises

We will use the `possum` dataset that we used in Precept 5.  Let's get ourselves familiar with the data.

```{r}
library(openintro)

?possum
dim(possum) # 104 rows and 8 columns

str(possum) # `possum` is a data frame

summary(possum)

```

## Different objects in R

**Two main classes of objects in R**:

* With homogeneous elements: Vectors, matrices, arrays.
* With heterogeneous elements: Lists, data frames.

## Different data types in R


* integer
* numeric
* logical
* character/string, 
* factor 
* complex (not covered in this course)

## Extractions in R

There are pretty much three main ways to extract elements out in a R object: with numerical indices, with names and with logical arguments.  The syntax changes slightly depending on the kind of R object that you are dealing with.

**Ex. Extract out the columns `site`, `age` from the dataset `possum` and look at the first 6 rows of the output.**

```{r}
head(possum) # `site` and `age` are the first and fourth columns

## Extraction with numerical indices
head(possum[, c(1, 4)])

## Extraction with names 
head(possum[, c("site", "age")])

```

**Ex. Extract out the rows for the possums with head length at least 100mm.**

```{r}
## Extraction with logical arguments
possum[possum$headL >= 100 , ]
```



## Calculate numerical summaries for each subset.

**Ex. What is the median total length (in cm) for the possums in Victoria and in Queensland/New South Wales, respectively?**
```{r}
tapply(X = possum$totalL, INDEX = possum$pop, FUN = median)
```

**Ex. Calculate the median total length (in cm) again for possums in Victoria and in Queensland/New South Wales for each gender group. **

```{r}
tapply(X = possum$totalL, INDEX = list(possum$pop, possum$sex), FUN = median)
```

Ex. Write a function that takes one of the four body measurements of the possums and returns the age(s) of the possum(s) that gives the maximum number of that body measurement.

```{r}
age.for.max = function(x){
  y = possum$age[x == max(x)]
  return(y)
}
  

age.for.max(x = possum$headL)
age.for.max(possum$skullW)
age.for.max(possum$totalL)
age.for.max(possum$tailL)

```


# Topics reviewed

* R object types
* Data types in R
* Extractions in R
* `tapply()` function
* How to create a function





